### 2010-08-31 Thesis/Chapter4.tex
- Update 26 August -
End chapter 5 + start of chapter 6 + bib updated + PhD comics added for later use :)



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@166 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 31 August -
Finished chapter 5 and 6 and started added references for chapter 7



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@169 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@

Finally, \cite{Miller07} proposed the total Lagrangian explicit dynamics (TLED) algorithm, a very efficient fully non-linear formulation. The algorithm is based on the finite element method using the total Lagrangian formulation, where stresses and strains are measured with respect to the original configuration. This choice allows for precomputing of most spatial derivatives before the commencement of the time-stepping procedure. The authors reported that the average number of floating-point operations per element per time step is $35\,$\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation. They were able to compute the deformation of a hexahedral mesh of $ 6\,000 $ elements in about $ 16\,$ms for a time step of $ 1\,$ms. If they could not achieve real-time computations, their work constituted a step towards the simulation of entire organs in real-time.

@@

\cite{Wu04} proposed the first application of GPU in medical simulation. They were able to achieve an interactive frame rate with topology changes in surgical simulation. Not only they used the condensation technique for merely calculating the surface nodes in the non-operation parts, but they implemented the conjugate gradient solver onto GPU. Indeed, entering 2003, the full IEEE floating point precision data type supported in shaders and texture stages allows the GPU to be applied into more general applications. The conjugate gradient solver includes three primary operations: multiplication of the sparse matrix and the vector, addition of two vectors and the sum-reduction operation. The core component is the multiplication of the sparse matrix and the vector. Therefore, they migrated this part of calculation from the CPU into the fragment processor of the GPU, to take advantage of the fragment processor in its efficient manipulation of the local texture memory on the mathematical calculation. The performance of their method on the GPU is up to $ 2.15 $ times faster than the CPU implementation.

>>>>

\cite{Mosegaard05a,Mosegaard05b} were probably the first to implement a mass-spring model on GPU in medical simulation. For the first time, the whole simulation of was performed on the GPU (physics, interaction and visualisation) and they used their framework in a congenital cardiac surgery simulation. They compared two implementations: the first one explicitly defined the connectivity between springs using a connectivity texture while the second one determines implicitly determines at runtime that springs are connected when they are neighbours. They were able to compute 188 iterations per second on a $42\,745$ particule model of a pig heart. Depending on the size of the model and the type of implementation, the authors achieved speed-ups between 10 and 30 $ \times $ over their CPU implementation.

Taking advantage of the GPU to reach a substantial speed-up is usually not straightforward. One must understand the limitations inherent in its design and devise algorithms accordingly. With this idea in mind, \cite{Sorensen06} reviewed the most important concepts and data structures required to realise two popular deformable models on the GPU: the spring-mass model and the finite element model.

\TODO{Other GPU-based mass-spring implementations?}

@@

In 2007, \citeauthor{Miller07} proposed the total Lagrangian explicit dynamics (TLED) algorithm, a very efficient fully non-linear formulation. The algorithm is based on the finite element method using the total Lagrangian formulation, where stresses and strains are measured with respect to the original configuration. This choice allows for precomputing of most spatial derivatives before the commencement of the time-stepping procedure. The authors reported that the average number of floating-point operations per element per time step is $35\, $\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation. They were able to compute the deformation of a hexahedral mesh of $ 6\,000 $ elements in about $ 16\,$ms for a time step of $ 1\,$ms. If they could not achieve real-time computations, their work constituted a step towards the simulation of entire organs in real-time.

Finally, \cite{Taylor07a} presented an efficient constitutive update procedure for viscoelastic models suitable for simulation of soft tissues at large strains and varying strain rates. The procedure is formulated for use in explicit dynamic finite element algorithms like the TLED algorithm. The procedure is based on a class of visco-hyperelastic constitutive models developed from purely hyperelastic strain energy functions by introducing relaxation terms and expressing in convolution integral form. The resulting constitutive equations are separated into rate-dependent and -independent terms, with the former being designated as state variables to be stored and updated at each time step also. These are converted to a differential form, which upon integration in time yields the required incremental update formula. They showed the validity of the procedure with a series of numerical examples.

@@

\cite{Wu04} proposed the first application of GPU in medical simulation \label{chap4:GPUMedicalSimulation}. They were able to achieve an interactive frame rate with topology changes in surgical simulation. Not only they used the condensation technique for merely calculating the surface nodes in the non-operation parts, but they implemented the conjugate gradient solver onto GPU. Indeed, entering 2003, the full IEEE floating point precision data type supported in shaders and texture stages allows the GPU to be applied into more general applications. The conjugate gradient solver includes three primary operations: multiplication of the sparse matrix and the vector, addition of two vectors and the sum-reduction operation. The core component is the multiplication of the sparse matrix and the vector. Therefore, they migrated this part of calculation from the CPU into the fragment processor of the GPU, to take advantage of the fragment processor in its efficient manipulation of the local texture memory on the mathematical calculation. The performance of their method on the GPU is up to $ 2.15 $ times faster than the CPU implementation.

### 2010-08-31 Thesis/Chapter5.tex
- Update 31 August -
Finished chapter 5 and 6 and started added references for chapter 7



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@169 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{The total Lagrangian explicit dynamics (TLED) algorithm}
\label{chap5}
\begin{shortAbstract}
TLED proposed by Miller almost real-time \\
GPU implementation to reach real-time \\
But first: TLED description \\
\OC{MICCAI 2008; MedIA 2009}
\end{shortAbstract}

@@

We adopt the notation of \citeauthor{Bathe95} with respect to indication of the relevant configuration of the body: a left superscript indicates the configuration in which a quantity occurs and, when applicable, a left subscript indicates the configuration with respect to which the quantity is measured. We note the coordinates of a point at time t $ ^t \mathbf{x}$.

@@

		\subsubsection*{Element nodal forces}
For a given element at time $ t $, the global nodal force contributions $ \mathbf{\tilde{F}} $ due to stresses in the element may be computed from:
\begin{equation}
\label{chap5:elementForce}
\leftidx{^t}{ \mathbf{\tilde{F}} }{} = \int_{^0 V} \, \leftidx{_0^t}{ \mathbf{B} }{_L^t} \, \leftidx{_0^t}{ \mathbf{\hat{S}} }{} \, d\leftidx{^0}{ V }{},
\end{equation}
where $ _0^t\mathbf{\hat{S}} $ is the vector form of the second Piola-Kirchhoff stress given by:
\begin{equation}
_0^t\mathbf{\hat{S}} = \left[ _0^t S_{11} \quad _0^t S_{22} \quad _0^t S_{33} \quad _0^t S_{12} \quad _0^t S_{23} \quad _0^t S_{13} \right] ^T,
\end{equation}
$ _0^t\mathbf{B}^T_L $ is the strain-displacement matrix and $ d^0V $ is the initial (undeformed) volume of the element. The strain-displacement matrix $ \mathbf{B}_L $ relates the strains in an element to the element's nodal displacements. If an implicit analysis is performed, the strain-displacement matrix is used in the assembly of the stiffness matrix. In explicit analyses it is used to compute element nodal force contributions according to \eqref{chap5:elementForce}. In geometrically linear analyses, we assume that the displacements are infinitesimally small so that the geometry of an element does not change over time and the strain-displacement matrix is constant. Conversely, because geometrically non-linear analyses (such as carried out by the TLED algorithm) take the change of geometry of the elements into account, the strain-displacement matrix $ \mathbf{B}_L = _0^t\mathbf{B}_L $ varies. However, the strain-displacement matrix at time $ t $ may be computed by transforming a linear matrix $ _0^t\mathbf{B}_{L0} $ using the deformation gradient. We begin by defining the linear matrix as:
\begin{equation}
_0^t\mathbf{B}_{L0} = \left[ _0\mathbf{B}_{L0}^{(1)} \quad _0\mathbf{B}_{L0}^{(2)} \quad \ldots \quad _0\mathbf{B}_{L0}^{(N)} \right],
\end{equation}
where $ N $ is the number of nodes per element and the submatrix $ _0\mathbf{B}_{L0}^{(a)}  $ are given by:
\begin{equation}
_0\mathbf{B}_{L0}^{(a)} =
	\begin{bmatrix}
	_0 h_{a,1} & 0 & 0 \\
	0 & _0 h_{a,2} & 0 \\
	0 & 0 & _0 h_{a,3} \\
	_0 h_{a,2} & _0 h_{a,1} & 0 \\
	0 & _0 h_{a,3} & _0 h_{a,2} \\
	_0 h_{a,3} & 0 & _0 h_{a,1}
	\end{bmatrix}
\quad a = 1, 2, \ldots, N.
\end{equation}
The subscripted comma denotes partial differentiation such as:
\begin{equation}
_0 h_{a,i} = \pd{h_a}{^0 x_i}.
\end{equation}
The full strain-displacement matrix (accounting for initial displacement effect) is then computed via
\begin{equation}
_0^t\mathbf{B}_{L}^{(a)} = \leftidx{_0^t}{ \mathbf{B} }{_{L0}^{(a)}} \, \leftidx{_0^t}{ \mathbf{X} }{^T}.
\end{equation} 
We note that the linear strain-displacement matrix is composed of derivatives of shape functions with respect to the original undeformed coordinates of the body. These derivatives are therefore constant and may be precomputed, affording the total Lagrangian formulation a significant computational advantage.

>>>>

\chapter{The total Lagrangian explicit dynamics (TLED) algorithm}
\label{chap5}
\begin{shortAbstract}
In 2007, \citeauthor{Miller07} proposed the total Lagrangian explicit dynamics (TLED) algorithm, a very efficient fully non-linear formulation. The authors reported that the average number of floating-point operations per element per time step is $35\, $\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation. Their work constituted a step towards the simulation of entire organs in real-time. Later in the year, \citeauthor{Taylor07b} reformulated the TLED algorithm to propose the first GPU implementation of a fully non-linear finite element method. However, restrictions due to using a graphics-based API necessitated reformulation of the force summation in their implementation. \cite{Taylor07a} also presented an efficient constitutive update procedure for viscoelastic models formulated for use in explicit dynamic finite element algorithms like the TLED algorithm. With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general-purpose GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the idea of a more straightforward and efficient implementation in mind. We also wanted to extend the simple non-linear TLED formulation by adding the constitutive update procedure for viscoelastic models introduced by \citeauthor{Taylor07a} and an anisotropic constitutive formulation as well. This work eventually yields a very efficient GPU-based non-linear, viscoelastic and anisotropic formulation for modelling solid organs in medical simulation and was the object of three publications \citep{Comas2008,Taylor2008,Taylor2009}. This chapter aims at describing the TLED algorithm in details, as well as introducing the viscoelasticity and anisotropy extensions. Their GPU implementations will be the object of the next chapter. 
\end{shortAbstract}

@@

We adopt the notation of \citeauthor{Bathe95} with respect to indication of the relevant configuration of the body: a left superscript indicates the configuration in which a quantity occurs and, when applicable, a left subscript indicates the configuration with respect to which the quantity is measured. We note the coordinates of a point $ ^t \mathbf{x}$ at time t.

@@

		\subsubsection*{Element nodal forces}
For a given element at time $ t $, the global nodal force contributions $ \mathbf{\tilde{F}} $ due to stresses in the element may be computed from:
\begin{equation}
\label{chap5:elementForce}
\leftidx{^t}{ \mathbf{\tilde{F}} }{} = \int_{^0 V} \, \leftidx{_0^t}{ \mathbf{B} }{_L^t} \, \leftidx{_0^t}{ \mathbf{\hat{S}} }{} \, d\leftidx{^0}{ V }{},
\end{equation}
where $ _0^t\mathbf{\hat{S}} $ is the vector form of the second Piola-Kirchhoff stress given by:
\begin{equation}
_0^t\mathbf{\hat{S}} = \left[ _0^t S_{11} \quad _0^t S_{22} \quad _0^t S_{33} \quad _0^t S_{12} \quad _0^t S_{23} \quad _0^t S_{13} \right] ^T,
\end{equation}
$ _0^t\mathbf{B}^T_L $ is the strain-displacement matrix and $ d^0V $ is the initial (undeformed) volume of the element. The strain-displacement matrix $ \mathbf{B}_L $ relates the strains in an element to the element's nodal displacements. If an implicit analysis is performed, the strain-displacement matrix is used in the assembly of the stiffness matrix. In explicit analyses it is used to compute element nodal force contributions according to \eqref{chap5:elementForce}. In geometrically linear analyses, we assume that the displacements are infinitesimally small so that the geometry of an element does not change over time and the strain-displacement matrix is constant. Conversely, because geometrically non-linear analyses (such as carried out by the TLED algorithm) take the change of geometry of the elements into account, the strain-displacement matrix $ \mathbf{B}_L = \leftidx{_0^t}{\mathbf{B}}{_L} $ varies. However, the strain-displacement matrix at time $ t $ may be computed by transforming a linear matrix $ _0^t\mathbf{B}_{L0} $ using the deformation gradient. We begin by defining the linear matrix as:
\begin{equation}
_0^t\mathbf{B}_{L0} = \left[ _0\mathbf{B}_{L0}^{(1)} \quad _0\mathbf{B}_{L0}^{(2)} \quad \ldots \quad _0\mathbf{B}_{L0}^{(N)} \right],
\end{equation}
where $ N $ is the number of nodes per element and the submatrix $ _0\mathbf{B}_{L0}^{(a)}  $ are given by:
\begin{equation}
_0\mathbf{B}_{L0}^{(a)} =
	\begin{bmatrix}
	_0 h_{a,1} & 0 & 0 \\
	0 & _0 h_{a,2} & 0 \\
	0 & 0 & _0 h_{a,3} \\
	_0 h_{a,2} & _0 h_{a,1} & 0 \\
	0 & _0 h_{a,3} & _0 h_{a,2} \\
	_0 h_{a,3} & 0 & _0 h_{a,1}
	\end{bmatrix}
\quad a = 1, 2, \ldots, N.
\end{equation}
The subscripted comma denotes partial differentiation such as:
\begin{equation}
_0 h_{a,i} = \pd{h_a}{^0 x_i}.
\end{equation}
The full strain-displacement matrix (accounting for initial displacement effect) is then computed via
\begin{equation}
_0^t\mathbf{B}_{L}^{(a)} = \leftidx{_0^t}{ \mathbf{B} }{_{L0}^{(a)}} \, \leftidx{_0^t}{ \mathbf{X} }{^T}.
\end{equation} 
We note that the linear strain-displacement matrix is composed of derivatives of shape functions with respect to the original undeformed coordinates of the body. These derivatives are therefore constant and may be precomputed, affording the total Lagrangian formulation a significant computational advantage.

### 2010-08-31 Thesis/Chapter6.tex
- Update 31 August -
Finished chapter 5 and 6 and started added references for chapter 7



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@169 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{GPU implementation of TLED}
\label{chap6}
\begin{shortAbstract}
\OC{ISBMS 2008a}
\end{shortAbstract}

@@

		\subsubsection*{Graphics API}
OpenGL (version 1.5 and newer) provides shader language based on the C programming language called OpenGL shading language (GLSL). In the DirectX API (DirectX 9 and newer), shaders are programmed with the high level shader language (HLSL). Cg is another high-level shading language developed by NVIDIA in close collaboration with Microsoft and it is very similar to Microsoft's HLSL. However, the Cg compiler is not specific to a graphics API and can output both OpenGL and DirectX shader programs.

@@

The Von Mises equivalent stress distributions resulting from stretching the model with and without inclusion of viscoelasticity are compared in \fig{chap6:fig-liver}. For the viscoelastic case (\fig{chap6:fig-liver}, right image) the model is depicted several seconds after the completion of the stretch, so that stresses had reached an approximately steady state. The stress dissipation introduced by the viscoelastic model is clearly evident, and for example would result in significantly altered virtual tool reaction forces.	
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter6/liver.pdf}
\end{center}
\caption[GPU solution times]{Deformation of a liver model using the TI model (middle) and the TIV model (right). The undeformed mesh is shown on the left. Colour maps indicate the relative Von Mises equivalent stress magnitude.} 
\label{chap6:fig-liver}
\end{figure}

>>>>

\chapter{GPU implementation of TLED}
\label{chap6}
\begin{shortAbstract}
The previous chapter introduced the TLED algorithm in details as well as enhancements with a viscoelastic and anisotropic formulation. If the TLED algorithm had already been implemented on GPU by \cite{Taylor07b}, restrictions due to using a graphics-based API and the limited hardware possibilities of GPUs necessitated reformulation of the force summation in their implementation. With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general computation on GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the aim to achieve a more straightforward and efficient implementation. The aim was also to enhance the GPU-based implementation of the TLED with a viscoelastic and anisotropic formulation. This chapter will describe the implementation of this non-linear, viscoelastic and anisotropic finite element method algorithm in the open source framework SOFA. By providing an efficient and accurate non-linear FEM for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators. This work has led to three publications \citep{Comas2008,Taylor2008,Taylor2009}.
\end{shortAbstract}

@@

		\subsubsection*{Graphics API}
OpenGL (version 1.5 and newer) provides shader language based on the C programming language called OpenGL shading language (GLSL). In the DirectX API (DirectX 9 and newer), shaders are programmed with the high level shader language (HLSL). Cg is another high-level shading language developed by NVIDIA in close collaboration with Microsoft and it is very similar to Microsoft's HLSL \citep{Mark03}. However, the Cg compiler is not specific to a graphics API and can output both OpenGL and DirectX shader programs.

@@

The Von Mises equivalent stress distributions resulting from stretching the model with and without inclusion of viscoelasticity are compared in \fig{chap6:fig-liver}. For the viscoelastic case (\fig{chap6:fig-liver}, right image) the model is depicted several seconds after the completion of the stretch, so that stresses had reached an approximately steady state. The stress dissipation introduced by the viscoelastic model is clearly evident, and for example would result in significantly altered virtual tool reaction forces.	
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter6/liver.pdf}
\end{center}
\caption[GPU solution times]{Deformation of a liver model using the TI model \textbf{(middle)} and the TIV model \textbf{(right)}. The undeformed mesh is shown on the left. Colour maps indicate the relative Von Mises equivalent stress magnitude.} 
\label{chap6:fig-liver}
\end{figure}
